---
title: "Jane Doe, Homework 1"
output:
  html_document:
    highlight: pygments
    theme: flatly
    toc: yes
---

Load packages you'll be using throughout the HW

```{r}
library(ggplot2)
```


Question 9
-----------------------------------------

### Q9 Part A

First, we calculated the mean of the non-normal distribution `D0`. 
  
```{r 9a_D0_mean}
x0 = c(1:3, 5, 7, 9)
p0 = c(.2, .4, .24, .10, .05, .01)
mu0 = sum(x0 * p0) / sum(p0)
print(mu0)
```

The mean of `D0` is 2.66.

**Instructor's Note:** You could also write your narrative text as follows, which will ensure that the number you report is always in sync with the code you wrote. This is a bit more complex, but recommended:

The mean of `D0` is `r mu0`.

### Q9 Part B 

Next, we drew a sample of 1000 observations from the skewed distribution `x0`.
  
```{r 9b_x0_sample}
o1 = sample(x0, 1000, replace=T, prob = p0)
```

Then, we made a histogram of `x0` along with the estimated probability density function.

```{r 9b_histogram, fig.width=7, fig.height=6}
df = data.frame(observations=o1)

ggplot(df, aes(x=observations)) +
  geom_histogram(aes(y=..density..), #..density.. normalizes the y axis to the density (i.e. counts / total)
                 color='black',
                 fill='white',
                 binwidth=1,
                 breaks=c(0.5:9.5)) + # see ?geom_histogram for details
  geom_density(alpha=0.2,
               fill='orange',
               adjust=3) +
  theme_bw()

## If you prefer plotting without ggplot, you can use this code:

# rs1 = hist(o1, prob=T, main='Sampling distribution of x0, 1000 Samples')
# lines(density(o1, adjust=3))
```
